---
title: "Wine Quality Prediction"
author: "Amine Agrane & Lydia Khelfane"
date : "10/02/2021" 
output: html_notebook
---


```{r echo=FALSE}
# import all the library
library(forcats)
library(ggplot2)
```

# Dataset and variables Presentation
In this notebook we will use the data from the Kaggle Repository (https://www.kaggle.com/rajyellow46/wine-quality) by P. Cortez, A. Cerdeira, F. Almeida, T. Matos, and J. Reis. The data include examples of red and white wines from Portugal-one of the world’s leading wine-producing countries. 

For each wine, a laboratory analysis measured characteristics such as acidity, sugar content, chlorides, sulfur, alcohol, pH, and density. The samples were then rated in a blind tasting by panels of no less than three judges on a quality scale ranging from zero (very bad) to 10 (excellent). In the case of judges disagreeing on the rating, the median value was used.


The wine dataset is composed by a total of 14 different variables : 

**1- fixed acidity (tartaric acid - g/dm3)** : most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

**2- volatile acidity (acetic acid - g/dm3)**: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

**3- citric acid (g/dm3) **: found in small quantities, citric acid can add ‘freshness’ and flavor to wines 

**4- residual sugar (g/dm3)** : the amount of sugar remaining after fermentation stops, it’s rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

**5- chlorides (sodium chloride - g/dm3) **: the amount of salt in the wine

**6- free sulfur dioxide (mg/dm3) **: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

**7- total sulfur dioxide (mg/dm3) **: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

**8- density (g/cm3)** : the density of water is close to that of water depending on the percent alcohol and sugar content

**9- pH** : describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

**10- sulphates (potassium sulphate - g/dm3)** : a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

**11- alcohol (% by volume)** : the percent alcohol content of the wine Output variable (based on sensory data):

**12- quality**: score between 0 and 10 that describes the quality of the wine.

**13- good**: (score between 0 and 10)

**14- color**: color of the wine. There are two type wine, red wine and white wine.

## Exploratory Data Analysis on the wine dataset
##### Load the wine dataset

We start by reading the data which is stored in the csv file `winequality.csv`. The file is loaded using the `read.csv` command, along with the `as.data.frame` command which store our data inside a structure of dataframe type.

```{r echo=TRUE}
# Load the data
df_wine <- as.data.frame( read.csv(file = './data/winequality.csv', sep=',', stringsAsFactors=F))
head(df_wine)
tail(df_wine)
```


#### Basic analysis on the dataset
Now we're gonna achieve some basic analysis on our wine dataset to get a better understanding of its contents, size and structure.

**Dimension of our wine Dataset :**
```{r echo=TRUE}
cat('The number of rows inside the dataset is : ', dim(df_wine)[1])
cat('The number of columns/features inside the dataset is : ', dim(df_wine)[2])
```

**Structure and distribution of the variables :**
  
We use the `str` and `summary` functions to display some basic statistics about our dataset. The `str` function shows the nature (type) of each variable (feature) of our dateset, and some values that the feature can take. The `summary` function in other hand give us some statistics metrics (min, max, median, etc) for each feature of the dataset.
```{r echo=TRUE}
str(df_wine)
summary(df_wine)
```
**Correlation between the variables :**
```{r}
as.data.frame( cor(df_wine[c(-13, -14)]))
```


**Number of NAN values for each variables :**
```{r}
cat('Number of NAN values for each variable :', colSums(is.na(df_wine)))
```



# Preparing the data
We want to predict the quality of the wine (score between 1 and 10) by using the rest of the feature variables. Here, wine quality is the variable we want to predict (Y variable).

First let's take a look at the distribution of the wine quality score in our dataset. We see that Wine quality appear to follow a bell shaped distribution (Gaussian/Normal distribution). This implies most wines are of average quality and few are good or bad
```{r}
# Simple Bar Plot
hist(df_wine$quality, main="Wine Quality Score Distribution",ylab="Count", xlab="Wine Quality Score", breaks=10, , col= "darkgreen")
```

Now let's check the frequency of each wine quality score. From the above table, we see that an approximate 45% of the wine quality scoes have a 6 score value.
```{r}
table(df_wine$quality)
```

## Binning the Quality Score Variable
We want to predict the quality of the wine based on the otehr 


We turn our Y variable into a two class variable. We will have the class "bad" for the lower wines (in terms of quality) and the class "good" for the best wines. We have to set a treshold on the wine quality score to assert if it's a good or bad wine.

First threshold :
- High : scores from 6 to 9.
- Low : scores from 1 to 5.


Scond threshold :
- High : scores from 6 to 9.
- Low : scores from 1 to 5.


```{r}
cat('First threshold class distribution : ', table(ifelse(df_wine$quality<=6,'Bad', ifelse(df_wine$quality>=7, "Good","")) ))
cat('Second threshold class distribution : ', table(ifelse(df_wine$quality<=5,'Bad', ifelse(df_wine$quality>=6, "Good",""))))
```

For a better class distribution, we'll choose low scores to be from 1 to 5 and high score from 6 to 9.


```{r}
qualityFactor <-ifelse(df_wine$quality<=4,'low', 
            ifelse(df_wine$quality>=5 & df_wine$quality<=7, "average", 
            ifelse(df_wine$quality>7,"high", "")))
wineData <- data.frame(df_wine, qualityFactor)
```


```{r}
ggplot(df_wine, aes(x =fct_reorder(qualityFactor, alcohol) , y = alcohol, fill=qualityFactor))+ geom_boxplot()
ggplot(df_wine, aes(x =fct_reorder(qualityFactor, density) , y = density, fill=qualityFactor))+ geom_boxplot()
```


## Data Preparation
As decision tree is the model used, there is no need for data preprocessing. Therefore, the training and test data can be prepared directly as the data is sorted randomly. A proportion of 75% to 25% is chosen. The partition of trainind and test dataset is done as below : 
```{r}
#training set
wine_train <- wine[1:3750, ]

#test set
wine_test <- wine[3751:4898, ]
```

