---
title: "Wine Quality Prediction"
author: "Amine Agrane & Lydia Khelfane"
date : "10/02/2021" 
output: html_notebook

---

```{r}
getwd()
```


# Exploratory Data Analysis on the wine dataset

##### Load the wine dataset
We start by reading the data which is stored in the csv file `winequality.csv`. The file is loaded using the `read.csv` command, along with the `as.data.frame` command which store our data inside a structure of dataframe type.

```{r}
# Load the data
wineData <- read.csv(file = './data/winequality.csv')
df_wine <- as.data.frame(wineData)
head(df_wine)
```
##### Basic analysis on the dataset
```{r}

```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

