---
title: "Wine Quality Prediction"
author: "Amine Agrane & Lydia Khelfane"
date: "10/02/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---


```{r echo=FALSE}
# import all the library
library(forcats)
library(ggplot2)
library(rpart)
library(rpart.plot)
```

# Project and Dataset Presentation    
In this notebook we will use the data from the Kaggle Repository (https://www.kaggle.com/rajyellow46/wine-quality) by P. Cortez, A. Cerdeira, F. Almeida, T. Matos, and J. Reis. The data include examples of red and white wines from Portugal-one of the world’s leading wine-producing countries. 

For each wine, a laboratory analysis measured characteristics such as acidity, sugar content, chlorides, sulfur, alcohol, pH, and density. The samples were then rated in a blind tasting by panels of no less than three judges on a quality scale ranging from zero (very bad) to 10 (excellent). In the case of judges disagreeing on the rating, the median value was used.

##### *Objective of this notebook*
The objective that we want to achieve throw this notebook is to build a machine learning model of classification type, that will predict if a wine is considered as good or note. The model takes as input some wine characteristics (alcohol content, acidity, sugar proportion, etc), and gives as output a binary variable that describes the quality of the wine ("Good" or "Bad"). We'll use a decision tree as our classification model.

##### *Dataset description*
The wine dataset is composed by a total of 14 different variables : 

**1- fixed acidity (tartaric acid - g/dm3)** : most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

**2- volatile acidity (acetic acid - g/dm3)**: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

**3- citric acid (g/dm3) **: found in small quantities, citric acid can add ‘freshness’ and flavor to wines 

**4- residual sugar (g/dm3)** : the amount of sugar remaining after fermentation stops, it’s rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

**5- chlorides (sodium chloride - g/dm3) **: the amount of salt in the wine

**6- free sulfur dioxide (mg/dm3) **: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

**7- total sulfur dioxide (mg/dm3) **: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

**8- density (g/cm3)** : the density of water is close to that of water depending on the percent alcohol and sugar content

**9- pH** : describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

**10- sulphates (potassium sulphate - g/dm3)** : a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

**11- alcohol (% by volume)** : the percent alcohol content of the wine Output variable (based on sensory data):

**12- quality**: score between 0 and 10 that describes the quality of the wine.

**13- good**: boolean variable related to the quality variable, is true when quality > 7 else the variable is false.

**14- color**: color of the wine. There are two type wine, red wine and white wine.

## Exploratory Data Analysis on the wine dataset
##### *Load the wine dataset*

We start by reading the data which is stored in the csv file `winequality.csv`. The file is loaded using the `read.csv` command, along with the `as.data.frame` command which store our data inside a structure of dataframe type.

```{r echo=TRUE}
# Load the data
df_wine <- as.data.frame( read.csv(file = './data/winequality.csv', sep=',', stringsAsFactors=F))
head(df_wine)
tail(df_wine)
```


## Basic analysis on the dataset
Now we're gonna achieve some basic analysis on our wine dataset to get a better understanding of its contents, size and structure.

**Dimension of our wine Dataset :**
```{r echo=TRUE}
cat('The number of rows inside the dataset is : ', dim(df_wine)[1])
cat('The number of columns/features inside the dataset is : ', dim(df_wine)[2])
```

**Structure and distribution of the variables :**
  
We use the `str` and `summary` functions to display some basic statistics about our dataset. The `str` function shows the nature (type) of each variable (feature) of our dateset, and some values that the feature can take. The `summary` function in other hand give us some statistics metrics (min, max, median, etc) for each feature of the dataset.
```{r echo=TRUE}
str(df_wine)
summary(df_wine)
```
**Correlation between the variables :**
```{r}
as.data.frame( cor(df_wine[c(-13, -14)]))
```


**Number of NAN values for each variables :**
```{r}
cat('Number of NAN values for each variable :', colSums(is.na(df_wine)))
```



# Preparing the data
We want to predict the quality of the wine (score between 1 and 10) by using the rest of the feature variables. Here, wine quality is the variable we want to predict (Y variable).

##### *Inspect the wine quality score feature*

Let's take a look at the distribution of the wine quality score in our dataset. We see that Wine quality appear to follow a bell shaped distribution (Gaussian/Normal distribution). This implies most wines are of average quality and few are good or bad

```{r}
# Simple Bar Plot
hist(df_wine$quality, main="Wine Quality Score Distribution",ylab="Count", xlab="Wine Quality Score", breaks=10, , col= "darkgreen")
```

Now let's check the frequency of each wine quality score. From the above table, we see that an approximate 45% of the wine quality scores have a 6 score value.

```{r}
table(df_wine$quality)
```



##### *Binning the Wine Quality Score Variable*

We want to predict the quality of the wine based on the rest of the variables present in our dataset. We see from the above table that the wine quality score variable takes 7 different values (3, 4, 5, 6, 7, 8, 9) inside our dataset. If we want to build a machine learning model to predict this variable, we'll use a classification model that will achieve a multiclass classification, in our case we have 7 distinct classes (7 possible values).

The question we want to answer through this notebook is whatever a wine is good or bad ? In our case, the output variable (Y) that we want to get is a binary variable 
True => "Good" and False => "Bad". 

We transform our win quality score variable to a binary variable by achieving a binning on its values, i.e we fix a specific threshold and associate each score to a unique class, the class "Bad" for the lower quality wines and the class "Good" for the best wines. 

Threshold :
  - Good : scores from 7 to 9.
  - Bad : scores from 1 to 6.


```{r}
cat('Threshold  distribution :', table(ifelse(df_wine$quality<=6,'Bad', ifelse(df_wine$quality>=7, "Good",""))))
```


```{r}
# We drop the quality column and replace it with the new qualityClass variable.
qualityClass <- as.factor(ifelse(df_wine$quality<=6,'Bad', ifelse(df_wine$quality>=7, 'Good','')))
df_wine <- data.frame(subset (df_wine, select = -quality), qualityClass)
```


# Build a Decision Tree Model
##### *Splitting the data into training and testing sets*
We split our data into two different sets : 

  - Training set for the model fitting (learning the patterns)
  - Testing set for estimating the model’s accuracy


```{r}
# we set the seed to get reproducible results
set.seed(10)

# split the datasets
dataset_size <- dim(df_wine)[1]
train_set_size <- round(0.8*dataset_size)
train_index <- sample(dataset_size, train_set_size, replace=FALSE)

training_data <- df_wine[train_index, ]
test_data <- df_wine[-train_index,]
```


##### *Instantiate and train the decision tree model*
Train the model the Using rpart library.

```{r}
rpart_model <- rpart(qualityClass~., data=training_data, method="class")
predict_rpart <- predict(rpart_model,  test_data[, -13], type="class")
mean(predict_rpart != test_data[, 13])
rpart.plot(rpart_model)
```

